---
title: "Getting Started with sleepwaker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sleepwaker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This guide demonstrates how to use **`sleepwaker`** to process ActiGraph batch sleep export data for behavioral research. The package provides automated workflows for both sleep analysis and physical activity research.

**What you'll learn:**

- How to install and load the package
- Three common usage scenarios
- Complete workflow from ActiLife export to analysis-ready data
- When to use each approach

# Installation

Install the development version from GitHub:
```{r install, eval=FALSE}
# Install sleepwaker
devtools::install_github("rdazadda/sleepwaker")
```

Load the package:
```{r load}
library(sleepwaker)
```

# Understanding the Data Format

## Input Requirements

**`sleepwaker`** requires batch sleep export data from ActiGraph ActiLife with three columns:

- **Subject Name**: Unique identifier for each participant
- **In Bed Time**: Datetime when participant went to bed (mm/dd/yyyy hh:mm:ss AM/PM)
- **Out Bed Time**: Datetime when participant got out of bed (mm/dd/yyyy hh:mm:ss AM/PM)

These columns are automatically included in ActiLife batch sleep exports.



## Sample Data

Let's load the real ActiGraph sample data included with the package:

```{r data, echo=TRUE, results='hide'}
# Load real sample data
data(sample.data)
```

```{r data-preview}
# Dataset summary
cat("Dataset contains", nrow(sample.data), "sleep periods for",
    length(unique(sample.data$`Subject Name`)), "participants")

# Showing all column names from ActiGraph export
cat("ActiGraph batch export includes these columns:\n")
cat(paste("-", names(sample.data), collapse = "\n"))
```

This is actual ActiGraph ActiLife batch sleep export data containing sleep periods detected by Cole-Kripke or Sadeh algorithms.

# Three Usage Scenarios

## Scenario 1: Format Sleep Periods for Sleep Analysis

Use case: You're conducting sleep research and need formatted sleep periods.
```{r scenario1}
sleep.periods <- sleepwaker(sample.data, 
                            input.type = "raw", 
                            output.type = "sleep")

# Display first few formatted sleep periods
head(sleep.periods, 5)
```

**What happened:**

- Batch export was reformatted into subject log diary format
- Timestamps split into separate date and time columns
- Data ready for sleep pattern analysis


## Scenario 2: Convert Sleep to Wake Periods for Activity Analysis

Use case: You're studying physical activity or sedentary behavior and need to exclude sleep.

```{r scenario2}
wake.periods <- sleepwaker(sample.data, 
                           input.type = "raw", 
                           output.type = "wake")

# Display first few wake periods
head(wake.periods, 5)
```

**What happened:**

- Sleep periods were inverted to create wake periods
- Each wake period represents time between consecutive sleep episodes
- Notice: Number of wake periods = number of sleep periods - 1 (n-1 per subject)

Why n-1 periods?: The final sleep period has no subsequent sleep to pair with, so no wake period can be generated after it.

## Scenario 3: Generate Both Sleep and Wake Periods

Use case: You need both sleep periods (for sleep analysis) and wake periods (for activity analysis).

```{r scenario3}
# Step 1: Create formatted sleep periods
sleep.periods <- sleepwaker(sample.data, 
                            input.type = "raw", 
                            output.type = "sleep")

# Step 2: Convert sleep periods to wake periods
wake.periods <- sleepwaker(sleep.periods, 
                           input.type = "sleep", 
                           output.type = "wake")

# Both datasets
cat("Sleep periods:", nrow(sleep.periods), "records\n")
cat("Wake periods:", nrow(wake.periods), "records\n")
```

**What happened:**

- First step: Formatted sleep periods for sleep analysis
- Second step: Generated wake periods from those sleep periods
- You now have both datasets for comprehensive analysis


# Saving Results to Files

All scenarios support direct export to CSV:
```{r save-examples, eval=FALSE}
# Save sleep periods
sleep.periods <- sleepwaker(sample.data, 
                            input.type = "raw", 
                            output.type = "sleep",
                            output.file = "sleep_periods.csv")

# Save wake periods  
wake.periods <- sleepwaker(sample.data, 
                           input.type = "raw", 
                           output.type = "wake",
                           output.file = "wake_periods.csv")
```

The function displays a confirmation message when files are saved successfully.
## Important Notes

**Supported:**

- Raw  to Sleep (formatting)

- Raw to Wake (formatting + inversion) 

- Sleep to Wake (inversion)

**Not supported:**

- Wake to Sleep (would cause data loss)


Why?: Once sleep periods are inverted to wake periods, the final sleep period is lost. Attempting to reconstruct it would require guessing, which would be scientifically invalid.
# Troubleshooting

## Common Issues

### Issue 1: Datetime Format Problems

Error message:
```
Error: Could not parse datetime
```

Solution: Check your data format first:
```{r validate-example, eval=FALSE}
validate.sleep.data(your_data, verbose = TRUE)
```

If parsing fails, standardize the format:
```{r fix-format, eval=FALSE}
your_data$`In Bed Time` <- standardize.datetime(your_data$`In Bed Time`)
your_data$`Out Bed Time` <- standardize.datetime(your_data$`Out Bed Time`)
```

The package handles these formats automatically:

- MM/DD/YYYY HH:MM:SS AM/PM (preferred)

- MM/DD/YYYY HH:MM:SS (24-hour)

- MM/DD/YYYY HH:MM AM/PM

- MM/DD/YYYY HH:MM

### Issue 2: Missing Required Columns

Error message:
```
Error: Missing required columns: In Bed Time
```

Solution: Check your column names match exactly:

- Subject Name

- In Bed Time

- Out Bed Time

Column names are case-sensitive.

### Issue 3: Insufficient Data Warning

Warning message:
```
Warning: Subject A001 has fewer than 2 periods. Skipping.
```

Explanation: Wake period calculation requires at least 2 sleep periods per subject. Subjects with only 1 sleep period are skipped.

Action: Review why the subject has limited data (short monitoring period, device removal, or algorithm failure).


# Getting Help

## Documentation

- **Function help**: Type `?sleepwaker` in R console
- **Package overview**: Type `help(package = "sleepwaker")`
- **Online reference**: [https://rdazadda.github.io/sleepwaker/](https://rdazadda.github.io/sleepwaker/)

## Support

- **Report bugs**: [GitHub Issues](https://github.com/rdazadda/sleepwaker/issues)
- **Ask questions**: [GitHub Discussions](https://github.com/rdazadda/sleepwaker/discussions)
- **Email**: rdazadda@alaska.edu

## Citation

If you use **`sleepwaker`** in your research, please cite:
```
Azadda, R.D., Grogan-Kaylor, A., & Lee, K. (2025). sleepwaker:
  Detect Sleep and Wake Periods from Accelerometry Data.
  R package version 0.1.0. https://github.com/rdazadda/sleepwaker
```

**Good luck with your research!**
